Kalman tracking and estimation of continuous gravitational waves with a pulsar timing array

### 2023-08-21 docs/manuscript/paper.tex
100 noise realisations, h5e-15, narrow f prior

\section{Signal tracking and parameter estimation} \label{sec:detect}
The set of static parameters $\boldsymbol{\theta}$ of the model outlined in Section \ref{sec:model} can be separated into parameters which correspond to the intrinsic frequency evolution of the pulsar and parameters of the GW source, i.e. 
\begin{equation}
	\boldsymbol{\theta} =  \boldsymbol{\theta}_{\rm psr} \cup \boldsymbol{\theta}_{\rm gw} \ , \label{eq:params1}
\end{equation}
with
\begin{equation}
	\boldsymbol{\theta}_{\rm psr} = \left \{ \gamma^{(n)}, f_{\rm em}^{(n)}(t_1),\dot{f}_{\rm em}^{(n)}(t_1),d^{(n)},\sigma^{(n)}\right\}_{1\leq n \leq N} \ ,
\end{equation}
and
\begin{equation}
	\boldsymbol{\theta}_{\rm gw} = \left \{h_0, \iota, \delta, \alpha, \psi, \Omega, \Phi_0 \right \} \ .  \label{eq:params3}
\end{equation}
In Equation \eqref{eq:params3} we have reparameterized the two GW polarisation strains, $h_{+}, h_{\times}$, in terms of the GW amplitude $h_0$ and the system inclination $\iota$:
\begin{align}
	h_+ &= h_0(1 + \cos^2 \iota) 	\label{eq:hphx} \ ,\\
	h_{\times} &= -2h_0\cos \iota 	\label{eq:hphx2} \ , 
\end{align}
where $\iota$ is the angle between the normal to the SMBHB orbital plane, $\boldsymbol{L}$, and the observer line of sight, i.e. $\cos \iota = \boldsymbol{n} \cdot \boldsymbol{L}$. We use this parametrisation in terms of $h_0$ and $\iota$ throughout the remainder of this work. For a PTA dataset containing $N$ pulsars we then have $7 + 5N$ parameters to estimate. Typically the pulsar parameters are better constrained than the GW parameters; for example estimates of pulsar distances are accurate to $\sim$ 10$\%$ \citep{Cordes2002astro.ph..7156C, Verbiest2012ApJ...755...39V, Desvignes2016,Yao2017}, but we have no prior information on the SMBHB GW source location.  \newline

@@

\begin{enumerate}
	\item Sample from the prior $\pi(\boldsymbol{\theta})$ to obtain $\pi(\boldsymbol{\theta})_{\rm sample}$
	\item Pass $\pi(\boldsymbol{\theta})_{\rm sample}$ to the Kalman filter
	\item Iterate over the data $\boldsymbol{Y}$ using the Kalman filter and obtain a single $\log \mathcal{L}$ value
	\item Return $\log \mathcal{L}$ to the nested sampler
	\item Nested sampler updates $\boldsymbol{\theta}$
	\item Repeat
\end{enumerate}
In Section \ref{sec:testing} we will apply this workflow on a synthetic data set.

@@

\subsubsection{Prior distributions}\label{sec:priors}

\textbf{CLARIFY re uninformative c.f loguniform}

\textbf{sin/cos priors realy uniform over cos(x)}

\textbf{referecne NANOgrav puslars and n index c.f. TAble ANTF}

\textbf{why did you choose NANOGrav you dick}

\textbf{priors on f, fdot, shrink to e.g. factor x ANTF error}

For the Bayesian methods used in this work, one must select reasonable priors, $\pi(\boldsymbol{\theta})$, for the complete set complete of static parameters. For $\pi(\boldsymbol{\theta}_{\rm gw})$ we choose standard non-informative priors \citep[e.g.][]{Bhagwat2021} as summarised in Table \ref{tab:parameters_and_priors}. The choice of $\pi(\boldsymbol{\theta}_{\rm psr})$ requires some additional discussion. \newline

@@

We can see from Figure \ref{fig:corner_plot_2} that the results from these 9 noise realisations agree well with the single realisation result that we saw in  Figure \ref{fig:corner_plot_1} for a single noise realisation. The estimates of the parameters $\Omega, \Phi_0, \psi, \delta$ and $\alpha$ are highly consistent across the different noise realisations with the injected value contained in the 90 \% credible interval for each of the marginalized posteriors. The variance between each of the individual 1D histograms is also generally low.

Conversely for the parameters $\iota$ and $h_0$ the variance between the different marginalized posteriors is generally high; we infer highly different estimates of these parameters depending on the particular realisation of the noise. Typically, for a given noise realisation, we do not contain the injected value within the 90\% credible interval for both $\iota$ and $h_0$ simultaneously.

@@

%\begin{figure*}
%	\centering
%	\subfigure[]{\includegraphics[width=0.24\textwidth]{example-image-a}} 
%	\subfigure[]{\includegraphics[width=0.24\textwidth]{example-image-a}} 
%	\subfigure[]{\includegraphics[width=0.24\textwidth]{example-image-a}}
%	\subfigure[]{\includegraphics[width=0.24\textwidth]{example-image-a}}
%	\caption{(a) blah (b) blah (c) blah (d) blah}
%	\label{fig:foobar}
%\end{figure*}

\section{Discussion}

Some potential further points for discussion

\begin{itemize}
	\item What is the performance of a single pulsar?
	\item How does this map to Hellings-downs?
	\item How does this map to having more than one source? Or a chrmomatic source?
	\item Non constant time sampling?
	\item Any further discussion on pulsar terms?
	\item PTAs like MSPs for small timing noise. Can we get away with large timing noise, and use more pulsars? Some pulsars are more useful than others, e.g. arXiv 2211.03201
	\item Can we also estimate radiometer noise? Is this useful?
	\item Lets preempt reviewer and make it explicit WHY we are not using TOAs
\end{itemize}

>>>>

\section{Signal tracking, parameter estimation and model selection} \label{sec:detect}
The set of static parameters $\boldsymbol{\theta}$ of the model outlined in Section \ref{sec:model} can be separated into parameters which correspond to the intrinsic frequency evolution of the pulsar and parameters of the GW source, i.e. 
\begin{equation}
	\boldsymbol{\theta} =  \boldsymbol{\theta}_{\rm psr} \cup \boldsymbol{\theta}_{\rm gw} \ , \label{eq:params1}
\end{equation}
with
\begin{equation}
	\boldsymbol{\theta}_{\rm psr} = \left \{ \gamma^{(n)}, f_{\rm em}^{(n)}(t_1),\dot{f}_{\rm em}^{(n)}(t_1),d^{(n)},\sigma^{(n)}\right\}_{1\leq n \leq N} \ ,
\end{equation}
and
\begin{equation}
	\boldsymbol{\theta}_{\rm gw} = \left \{h_0, \iota, \delta, \alpha, \psi, \Omega, \Phi_0 \right \} \ .  \label{eq:params3}
\end{equation}
In Equation \eqref{eq:params3} we have reparameterized the two GW polarisation strains, $h_{+}, h_{\times}$, in terms of the GW amplitude $h_0$ and the system inclination $\iota$:
\begin{align}
	h_+ &= h_0(1 + \cos^2 \iota) 	\label{eq:hphx} \ ,\\
	h_{\times} &= -2h_0\cos \iota 	\label{eq:hphx2} \ , 
\end{align}
where $\iota$ is the angle between the normal to the SMBHB orbital plane, $\boldsymbol{L}$, and the observer line of sight, i.e. $\cos \iota = \boldsymbol{n} \cdot \boldsymbol{L}$. We use this parametrisation in terms of $h_0$ and $\iota$ throughout the remainder of this work. For a PTA dataset containing $N$ pulsars we then have $7 + 5N$ parameters to estimate. Typically the pulsar parameters are better constrained than the GW parameters; for example estimates of pulsar distances are accurate to $\sim$ 10$\%$ \citep{Cordes2002astro.ph..7156C, Verbiest2012ApJ...755...39V, Desvignes2016,Yao2017}, but we have no prior information on the SMBHB GW source location.  \newline

@@

\begin{enumerate}
	\item Sample from the prior $\pi(\boldsymbol{\theta})$ to obtain $\boldsymbol{\theta}_{\rm sample}$
	\item Pass $\boldsymbol{\theta}_{\rm sample}$ to the Kalman filter
	\item Iterate over the data $\boldsymbol{Y}$ using the Kalman filter and obtain a single $\log \mathcal{L}$ value
	\item Return $\log \mathcal{L}$ to the nested sampler
	\item Nested sampler updates $\boldsymbol{\theta}$
	\item Repeat
\end{enumerate}
In Section \ref{sec:testing} we will apply this workflow on a synthetic data set.

@@

\subsubsection{Prior distributions}\label{sec:priors}
For the Bayesian methods used in this work, one must select reasonable priors, $\pi(\boldsymbol{\theta})$, for the complete set complete of static parameters. For $\pi(\boldsymbol{\theta}_{\rm gw})$ we choose standard non-informative priors \citep[e.g.][]{Bhagwat2021} as summarised in Table \ref{tab:parameters_and_priors}. The choice of $\pi(\boldsymbol{\theta}_{\rm psr})$ requires some additional discussion. \newline

@@

We can see from Figure \ref{fig:corner_plot_2} that the results from these 9 noise realisations agree well with the single realisation result that we saw in  Figure \ref{fig:corner_plot_1} for a single noise realisation. The estimates of the parameters $\Omega, \Phi_0, \psi, \delta$ and $\alpha$ are highly consistent across the different noise realisations with the injected value contained in the 90 \% credible interval for each of the marginalized posteriors. The variance between each of the individual 1D histograms is also generally low. Conversely for the parameters $\iota$ and $h_0$ the variance between the different marginalized posteriors is generally high; we infer highly different estimates of these parameters depending on the particular realisation of the noise. Typically, for a given noise realisation, we do not contain the injected value within the 90\% credible interval for both $\iota$ and $h_0$ simultaneously. \newline

@@

\section{Discussion}
We have shown how our complementary approach to PTA data analysis using a Kalman filter can successfully recover the underlying parameters of the model and calculate Bayes factors in order to perform Bayesian model selection.

The natural first extension of this work is to implement the method whilst retaining the pulsar terms of Equation \ref{eq:measurement}.

Whilst PTA searches commonly drop the pulsar terms as standard practice, we have shown that this leads to biases in the inferred 1D marginalized posteriors for the model parameters.

Biases in synthetic datasets from dropping the pulsar terms are also observed by \cite{Zhupulsarterms} and \cite{Chen2022}. Alternatively, if one wishes to continue to use just the Earth terms it would be highly desirable to

We have considered only a specific configuration of pulsars composing our synthetic PTA - the same pulsars that make up NANOGrav. Whilst our method is generally independent of the specific choice of pulsars, it would be of interest to compare the performance of the method using different pulsars configurations and different PTAs such as PPTA and the EPTA. Indeed, due to the increased computational demands as pulsars are added to the arrays it may be advantageous to optimally select a subset of the total number of available pulsars which may be able to provide comparable performance to a larger PTA  \citep{2023MNRAS.518.1802S}.

Throughout this work we have assumed that all pulsars are observed at the same time, with a constant time sampling. In practice this assumption does not hold and pulsars will be observed with different cadences at different time intervals. Applying this method to non-constant time sampling is a key further extension.

We have also assumed that there is only a single, quasi-monochromatic GW source that influences the pulsar frequencies.However, it may be possible to resolve multiple continuous GW source simultaneously. Our method naturally extends to multiple GW sources; Equation \ref{eq:measurement} is modified to become the linear superposition of multiple GWs. The stochastic background itself is the incoherent summation of many individual GW sources. It therefore stands to reason that we could extend this method to also search for a stochastic GW background without requiring cross-correlating the pulsar residuals to uncover the Hellings-Downs curve.

\textcolor{red}{TK: Lets pre-empt reviewer and make it explicit WHY we are not using TOAs}
We have formulated the state space model in terms of frequencies, where
